---
title: "Who were the best Premier League Players in their positions"
output: html_document
date: "2025-11-10"
author: " By Theodore Etuwewe & Usman Agboola" 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(readr)
library(scales)
library(stringr)
premleague <- read.csv("premleague.csv")
```

## *Introduction*
The Premier League is a battlefield, and every player is fighting for glory. Because of this, we were inspired to find out who was truly number 1 in their positions. The analysis aims to dive into the stats that separate the stars from the rest; goals, assists, tackles, interceptions, clean sheets, and more. By looking at offensive firepower, defensive strength, and goalkeeping feats, we aim to crown the ultimate player of the season, the one who dominates all players of the game.

We got this dataset from Kaggle, and here's a link to it: [English Premier League Player Stats](https://www.kaggle.com/datasets/rishikeshkanabar/premier-league-player-statistics-updated-daily)

But before we look into the data, lets clean it up a bit. 

# *Cleaning Up process*

```{r}

premleague <- premleague %>%
  mutate(across(everything(), ~ ifelse(is.na(.), 0, .))) %>%  # changes NA values to zero
  filter(!is.na(Name) & Name != "") %>%                       # removes rows with missing player name     
  distinct()                                                  # removes duplicate rows

glimpse(premleague)
```

As you can see, there is a lot to look at here. This dataset contains detailed statistics for Premier League players like their offensive metrics, defensive metrics, and goalkeeper-specific statistics. What makes it interesting is the depth of information for each player, allowing analysis of scoring efficiency, defensive performance, and overall contribution to the team. It enables comparisons between players across positions, highlighting strengths and patterns that are not obvious from simple totals, such as who is the most efficient scorer or the most influential defender.




# *What do we intend to learn from this data?*

# *Core question*
1. Who is the best person to ever touch the Field? 

# *Other questions*
1. Who scored the most goals?

2. Who provided the most assists?

3. Who contributes the most per appearance?

4. How does each position contribute to scoring goals or performing assists?

5. Does age affect a players involvement in goals or assists?

6. Which clubs had the most productive players?

7. What percentage of a teams goals does each player contribute?


*Alright*. enough chitchat. I believe its time to get down and dirty with the details.

# *1.Who scored the most goals?*

```{r}
topgoals <- premleague %>% 
  arrange(-Goals) %>% 
  slice_head(n = 10)

ggplot(topgoals, aes(x = reorder(Name, Goals), y = Goals)) +
  geom_col(fill = "steelblue") + 
  geom_text(aes(label = Goals), hjust = -0.1) +
  coord_flip() +
  labs(
    title = "Top 10 Goal Scorers",
    x = "Player Name" ,
    y = "Goals") +
theme_minimal()
```


# *2. Who provided the most assists?*

```{r}
topassists <- premleague %>% 
  arrange(-Assists) %>% 
  slice_head(n = 10)

ggplot(topassists, aes(x = reorder(Name, Assists), y = Assists, fill = Assists)) +
  geom_col() +
  geom_text(aes(label = Assists), hjust = -0.1) +
  labs(
    title = "Top 10 Assist Makers",
    x = "Player" ,
    y = "Assists") +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +
  theme_minimal() +
  coord_flip()
```



# *3. Who contributes the most per appearance?* 

```{r}

premleague <- premleague %>%
  mutate(GoalInvolvement = Goals + Assists,
         InvolvementPerAppearance = GoalInvolvement / Appearances)

top10efficiency <- premleague %>%
  filter(Appearances > 5) %>%
  arrange(-InvolvementPerAppearance) %>%
  slice_head(n = 10)

ggplot(top10efficiency, aes(x = InvolvementPerAppearance, 
                             y = reorder(Name, InvolvementPerAppearance))) +
  geom_segment(aes(x = 0, 
                   xend = InvolvementPerAppearance, 
                   yend = Name), 
               color = "grey", linewidth = 1.2) +
  geom_point(color = "darkblue", size = 4) +
  geom_text(aes(label = round(InvolvementPerAppearance, 2)), 
            hjust = -0.3, size = 3.5) +
  labs(
    title = "Top 10 Players by Goal Involvement per Appearance",
    x = "Goal Involvement per Appearance",
    y = "Player"
  ) +
  theme_classic() 

```


# *4. How does each position contribute to scoring goals or performing assists?*

```{r}
ggplot(premleague, aes (x = Position, y = Goals + Assists, fill = Position)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Goal Involvement by Position",
    x = "Postion",
    y = "Goal Involvement (Goals & Assists)"
  ) + 
  theme_classic()+
  theme(legend.position = "none")
```


# *5. Does age affect a players involvement in goals or assists?*

From old legends like Thiago Silva to young new faces like Trent Alexander Arnold, we aim to find out if father time really does have a hand in the performance of a player. 

```{r}
ggplot(premleague, aes(x = Age, y = Goals + Assists, color = Age)) +
  geom_point(alpha = 0.7, size = 4) +
  scale_color_gradient(low = "steelblue", high = "red") +
  labs(
    title = "Age vs Goal/Assist Involvement",
    x = "Age",
    y = "Goal Involvement (Goals + Assists)",
    color = "Age"
  ) +
theme_minimal()
```

As you can see, father time unfortunately does. Younger players are more likely to be bodied out by their older counterparts and the older players simply may not have the stamina to keep up with the younger ones. But it would seem that those between the ages of 25-30 seem to be in peak physical condition, so they don't have much to worry about. 


# *6. Which clubs had the most productive players?*

7. What percentage of a teams goals does each player contribute?

